Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-29 _tex/paper2.tex
Built site for gh-pages

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study discusses two prominent issues
that can compromise the reliability and validity of CJ's trait
estimates. The first issue concerns the overreliance on Thurstone's Case
V assumptions in the statistical analysis of CJ data. The second
involves the apparent disconnect between CJ's approach to trait
measurement and hypothesis testing. The study then addresses these
challenges by adopting a more systematic and integrated approach
grounded in causal inference and Bayesian inference methods.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} updates the Thurstonian model
to address these challenges. The model integrates the theoretical core
principles alongside key assessment design features relevant to CJ
experiments, such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model for the analysis of
pairwise comparison data. Section~\ref{sec-discussion} discusses the
implications of the findings and explores avenues for future research.
Finally, Section~\ref{sec-conclusion} summarizes the study's key
insights.

@@

This section discusses the two prominent issues that can undermine the
reliability and validity of CJ's trait estimates.
Section~\ref{sec-theory-issue1} examines the overreliance on Thurstone's
Case V assumptions in the statistical analysis of CJ data.
Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing.

@@

This section addresses the challenges outlined in
Section~\ref{sec-theory-issues} by leveraging the structural approach to
causal inference to extend Thurstone's model. The extension combines the
core theoretical principles from Section~\ref{sec-thurstone_theory} with
several key assessment design features relevant to CJ experiments, such
as the selection of judges, stimuli, and comparisons. Specifically,
Section~\ref{sec-theory-theoretical_P} incorporates the theoretical
principles into a \emph{conceptual-population model}. Conversely,
Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
features into a \emph{sample-comparison model}. Combining these elements
helps to elucidate how all actors and processes interact during CJ
experiments.

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. This approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ experiments. Essentially, SCMs and
DAGs function as \emph{conceptual models} on which identification
analysis rests \citep[pp.~4]{Schuessler_et_al_2023}. Identification
analysis helps researchers determine whether an estimator can accurately
compute an estimand based solely on its (causal) assumptions, regardless
of random variability \citep[pp.~4]{Schuessler_et_al_2023}. Here,
\emph{estimands} represent the specific quantities researchers aim to
determine \citep{Everitt_et_al_2010}. \emph{Estimators} denote the
methods or functions that transform data into an estimate
\citep{Neal_2020}, while \emph{estimates} are the numerical values
approximating the estimand. These estimates are obtained through
\emph{estimation}, which is the process that integrates data with an
estimator \citep{Everitt_et_al_2010}.

For instance, researchers might aim to answer the question: ``To what
extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, they could
design a CJ experiment by randomly assigning students to two groups,
each exposed to a different teaching method. Afterward, judges would
compare pairs of written texts, generating a dichotomous outcome that
reflects the relative quality of each text. In this scenario,
researchers can rephrase the research question as the estimand:
``\emph{On average}, is there a difference in the ability to produce
high-quality written texts between the two groups of students?''.
Researchers would then use estimates from the BTL model, or its
transformations, to approximate this estimand.

However, Section~\ref{sec-theory-issues} provides compelling evidence
that the BTL model suffers from several statistical and measurement
issues, including identification problems. These issues ultimately raise
concerns about the consistency of the model's estimates. An estimator is
\emph{consistent} if it converges to the ``true'' value of an estimand
as the data size approaches infinity \citep{Everitt_et_al_2010}. Since
identification is a prerequisite for consistency, the absence of proper
identification means that researchers cannot achieve consistency, even
with ``infinite'' and error-free data. As a result, deriving meaningful
insights from finite data also becomes impossible
\citep[pp.~5]{Schuessler_et_al_2023}.

Luckily, SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in an interactive, non-parametric manner. This flexibility
enables feasible identification strategies without defining the
variables' data types, the functional forms, or their parameters
\citep[pp.~35]{Pearl_et_al_2016}.

@@

Before integrating the theoretical principles from
Section~\ref{sec-thurstone_theory}, it is important to define SCMs and
DAGs. SCMs are formal mathematical models defined by a set of
\emph{endogenous} variables, a set of \emph{exogenous} variables, and a
set of functions \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These function use the
symbol `\(:=\)' to denote the asymmetrical causal dependence between
variables and the symbol `\(\:\bot\:\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

@@

Figure~\ref{fig-CJ1} displays CJ's outcome, discriminal difference,
discriminal processes and judges' biases assumed by the Thurstone's
theory. Notice that in the figure, the structural

>>>>

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues that can undermine the reliability and validity of CJ's
trait estimates. The first issue arises from the excessive reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
The second stems from the apparent disconnect between CJ's approach to
trait measurement and hypothesis testing. The study then addresses these
issues by extending Thurstone's model through a systematic and
integrated approach based on causal and Bayesian inference methods.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's model to
address these challenges. The model integrates the theoretical core
principles alongside key assessment design features relevant to CJ
experiments, such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model for the analysis of
pairwise comparison data. Section~\ref{sec-discussion} discusses the
implications of the findings and explores avenues for future research.
Finally, Section~\ref{sec-conclusion} summarizes the study's key
insights.

@@

This section identifies and discusses the two prominent issues that can
undermine the reliability and validity of CJ's trait estimates.
Section~\ref{sec-theory-issue1} examines the the excessive reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
Section~\ref{sec-theory-issue2} focuses on the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing.

@@

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. The approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ experiments.

Essentially, SCMs and DAGs function as \emph{conceptual models} on which
identification analysis rests \citep[pp.~4]{Schuessler_et_al_2023}.
\emph{Identification analysis} helps researchers to determine whether an
estimator can accurately compute an estimand based solely on its
(causal) assumptions, regardless of random variability
\citep[pp.~4]{Schuessler_et_al_2023}. Here, \emph{estimands} represent
the specific quantities researchers aim to determine
\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
functions that transform data into an estimate, while \emph{estimates}
are the numerical values approximating the estimand
\citep{Neal_2020, Everitt_et_al_2010}.

For instance, researchers might aim to answer the question: ``To what
extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, they could
design a CJ experiment by randomly assigning students to two groups,
each receiving a different teaching method. Judges would then compare
pairs of students' written texts, producing a dichotomous outcome that
reflects the relative quality of each text. In this scenario,
researchers can reformulate the research question as the estimand:
``\emph{On average}, is there a difference in the ability to produce
high-quality written texts between the two groups of students?''.
Researchers would then use estimates from the BTL model, or its
transformations, to approximate this estimand.

However, Section~\ref{sec-theory-issues} presents strong evidence that
the BTL model has several statistical and measurement limitations. These
limitations hinder the model's ability to identify distinct estimands
relevant to CJ inquiries, including the one presented in the previous
paragraph. Identification is crucial because it is a necessary condition
for ensuring consistent estimators. Without identification, researchers
cannot achieve consistency even with ``infinite'' and error-free data.
In other words, the estimates derived from the estimator fail to
converge to the ``true'' value of the estimand as the data size
approaches infinity \citep{Everitt_et_al_2010}. As a result, deriving
meaningful insights from finite data also becomes unattainable
\citep[pp.~5]{Schuessler_et_al_2023}.

Luckily, SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in an interactive, non-parametric way. This flexibility
enables feasible identification strategies without requiring
specification of the types of variables, the functional forms relating
them, or the parameters of those functional forms
\citep[pp.~35]{Pearl_et_al_2016}.

As a result, this section addresses the identified limitations by
extending Thurstone's model through a systematic and integrated approach
based on the structural approach to causal inference. Specifically, it
leverages the capabilities of this approach to formalize the combination
of the core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments. These features include the selection of
judges, stimuli, and comparisons. Combining these components clarifies
the interactions among all actors and processes involved in CJ
experiments. Accordingly, Section~\ref{sec-theory-theoretical_P}
incorporates the theoretical principles into a
\emph{conceptual-population model}. Meanwhile,
Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
features into a \emph{sample-comparison model}.

@@

Before integrating the theoretical principles from
Section~\ref{sec-thurstone_theory}, it is important to define SCMs and
DAGs. SCMs are formal mathematical models defined by a set of
\emph{endogenous} variables, a set of \emph{exogenous} variables, and a
set of functions \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables.

For instance, Figure~\ref{fig-CJ1_scm} display the SCM relating the CJ
outcome \((O_{kjabi_{1}i_{2}})\), discriminal difference
\((D_{kjabi_{1}i_{2}})\), perceived discriminal processes for each text
within each student \((PT_{kjai_{1}}, PT_{kjbi_{2}})\), the ``true''
discriminal processes for each text within each student
\((T_{kjai_{1}}, T_{kjbi_{2}})\), and the judges' biases \((B_{kj})\).
Notice that the structural equations use the symbol `\(:=\)' to denote
the asymmetrical causal dependence between variables and the symbol
`\(\:\bot\:\)' to represent \emph{d-separation}, a concept akin to
(conditional) independence.

@@
